---
title: "Tutorial of BayeSMART on DLPFC Data"
author: "Yanghong Guo"
date: "2024-08-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```



## BayeSMART on all 12 slides of three donors

```{r}

load("/Users/yanghongguo/Desktop/Research/iBurst/data/4 dataset final/DLPFC_p1.RData")
load("/Users/yanghongguo/Desktop/Research/iBurst/data/4 dataset final/DLPFC_p2.RData")
load("/Users/yanghongguo/Desktop/Research/iBurst/data/4 dataset final/DLPFC_p3.RData")

cntm <- c(cnts_DLPFC_p1, cnts_DLPFC_p2, cnts_DLPFC_p3)
xys <- c(xys_DLPFC_p1, xys_DLPFC_p2, xys_DLPFC_p3)

# We get the matrix that contains all the gene count matrices
layers = matrix(NA, nrow = 33538, ncol = 0)
each_length <- c()

for (s in 1:12){
  layer1 <- cntm[[s]]
  layers <- cbind(layers, layer1)
  each_length <- c(each_length, ncol(layer1))
}

```

Since there is no such deep learning model to analyze the HD-stained images of DLPFC, we use the reference-free method STDeconvolve instead to get the image profile.
```{r}
library(STdeconvolve)

counts <- cleanCounts(layers, min.lib.size = 1)
corpus <- restrictCorpus(counts, removeAbove=1.0, removeBelow = 0.05)
ldas <- fitLDA(t(as.matrix(corpus)), Ks = c(5,6,7,8,9,10))
## get best model results
optLDA <- optimalModel(models = ldas, opt = "min")
## extract deconvolved cell-type proportions (theta) and transcriptional profiles (beta)
results <- getBetaTheta(optLDA, perc.filt = 0.05, betaScale = 1000)
deconProp <- results$theta

V <- round(as.matrix(deconProp)*100)
```

Then get the geospatial profile $G$ and molecular profile $Y$.
```{r}
G <- matrix(ncol = 6, nrow = 0)
G_origin <- NULL
spot_xys <- NULL
list_count <- NULL

for (s in 1:12){

  M1 <- as.matrix(cntm[[s]])
  
  spot_xy <- xym[[s]]
  G1 <- get.neighbor(spot_xy, 6)

  list_count[[s]] <- M1
  G_origin[[s]] <- G1
  
  if (s > 1){
    non_zero_indices <- G1 != 0
    G1[non_zero_indices] <- G1[non_zero_indices] + sum(each_length[1:(s-1)])
  }
  
  # store the matrix
  G <- rbind(G, G1)
  
  spot_xys[[s]] <- spot_xy
  print(paste0("Done with sample ", s))
}

Y <- batch_remove(list_count, xys = xys, gene_select = "sparkx", n_gene = 2000, pcn = 3)

```

Note that up to this point, the geospatial profile $G$ only contains neighborhood information within each section, not across adjacent sections. To include the neighborhood information between adjacent slides, we need to add two more columns to the matrix $G$ manually by assigning spots with the same x and y axis values as neighbors from two adjacent slides. No neighborhood information between slides from different individuals.

```{r}
xys_1 <- lapply(xys_DLPFC_p1, function(info.i){
  as.matrix(info.i[, c("x", "y")])
})
xys_2 <- lapply(xys_DLPFC_p2, function(info.i){
  as.matrix(info.i[, c("x", "y")])
})
xys_3 <- lapply(xys_DLPFC_p3, function(info.i){
  as.matrix(info.i[, c("x", "y")])
})

G_extra_1 <- G_extra(xys_1)
G_extra_2 <- G_extra(xys_2)
G_extra_3 <- G_extra(xys_3)

non_zero_indices <- G_extra_2 != 0
G_extra_2[non_zero_indices] <- G_extra_2[non_zero_indices] + nrow(G_extra_1)

non_zero_indices <- G_extra_3 != 0
G_extra_3[non_zero_indices] <- G_extra_3[non_zero_indices] + nrow(G_extra_1) + nrow(G_extra_2)

G_extra <- rbind(G_extra_1, G_extra_2, G_extra_3)

G <- cbind(G, G_extra)
```

Then run BayeSMART on all 12 slides.
```{r}
K <- 7
w <- 0.1

result <- run.BayeSMART(V, Y, G, n_cluster = K, w = w)

spatial_domain <- get.spatial.domain(result)
domain_res <- domain_split(spatial_domain, G_origin)
```

With the clustering result we can calculate the ARI for each slide.
```{r}
annotations <- lapply(xys, function(info.i){
  as.matrix(info.i[, c("annotation")])
})

# calculate ARI
library(mclust)
library(aricode)
scores_ARI <- c()
for (s in 1:12){
  annotation <- annotations[[s]]
  domains <- domain_res[[s]]
  
  ari_score <- adjustedRandIndex(domains, annotation)
  scores_ARI <- c(scores_ARI, ari_score)
}

scores_ARI
mean(scores_ARI)
```